Living on the edge: crop boundary effects on canola and wheat yields in Alberta

### 2021-09-08 Manuscript/manuscript.Rmd
Worked on discussion

datSourceFinal <- datSource %>% filter(use) %>% filter(grepl('Wheat|Canola',crop)) 
yearsPerField <- datSourceFinal %>% count(field) %>% pull(n) #How many years of data at each unique location?

#Boundary lengths
bLengths <- read.csv('../Data/boundaryLengths.csv')
bNames <- c('Bare','Grassland','Other Crop','Shelterbelt','Standard','Wetland')

@@

use <- with(datSource,use&modelComplete2&crop=='Wheat')
wheatMods <- lapply(paths[use],function(x) getModelInfo(x)$smooths) %>% 
  set_names(nm=datSource$filename[use]) %>% 
  bind_rows(.id='field') %>% 
  mutate(pval=as.numeric(pval)) %>% 
  group_by(Smoother) %>% 
  summarize(meanP=mean(pval),maxP=max(pval),propSig=sum(pval<0.05)/n(),meanEDF=mean(edf),meanRef=mean(Ref.df))
rm(use,paths)

@@

Intensive agricultural production has increased over the last 100 years, and agricultural land now makes up over a third of ice-free land on Earth
This has allowed increases in human population and increased (global) stability in production
However, this is not without cost, as higher-diversity non-crops are converted to lower-diversity crops, resulting in loss of habitat and overall biodiversity of non-target organisms
Maintaining both biodiversity and production in agroecosystems represents a seldom-considered goal of conservationists and agronomists, and hold the potential for win-win scenarios
Key to this is the preservation of semi-natural land (SNL), which represents the interface between crops and non-crops within agroecosystems

SNL in and around crops is important for both agricultural production and conservation
They are habitat for mobile organisms, and can therefore act as sources of ecosystem services such as pollination or pest control (@albrecht2020; @gardner2021)
They also can create microclimate effects that reduce extreme temperature, trap moisture, and reduce wind speed (@kort1988)
Unfortunately, most of the research is concentrated in Europe, and they tend to be less-studied on other continents
In particular, North American agroecosystems have larger fields, different crop varieties, and agronomic practices, all of which could negate or confound potential effects of SNL

<!-- - SNL has conflicting effects on crop yield because of ecosystem service provision as well as negative edge effects -->
SNL may affect yields at intermediate distances, depending on the spatial scale at which ecosystem services operate.
Edge effects cause low yields at the edge of crops because of sparse or late seedling emergence, poor microclimate, and competition with weeds.
At the same time, the centre of large fields will not receive ecosystem services if they decay with distance from edges (@kowalchuk1995).
For example, pollination services from central place foragers that nest in SNL but forage in crops drops rapidly with distance.
Therefore, yield may be maximized at intermediate distances, where the ecosystem services cancel out negative edge effects (e.g. @vanVooren2017).
This suggests a ``goldilocks" field size, where negative edge effects are canceled out by ecosystem services.
<!-- - Less-considered are ecosystem ``disservices", where weeds and pests  -->
<!-- - In North America, there are financial incentives to make fields large, homogeneous, and with simple geometry -->

```{r rawData, echo=FALSE, fig.cap='Potential yield patterns, depending on ecosystem service effects with distance. In both figures, negative environmental effects (edge effects) are shown in red, and decrease with distance from the edge of the crop, creating an increase in yield with distance. The left figure shows a decline in ecosystem services with distance, leading to an intermediate peak in yield, while the right figure shows an increase in services with distance, leading to a peak at the centre of the field (note: this also occurs if ecosystem services do not depend on distance).', out.width='75%'}  
  include_graphics('../Figures/ExamplePlots/hypotheses.png',dpi=NA)
```

Studies of SNL effects on crop yield also suffer from limited scope (e.g. few crop types) and small sample sizes, limiting inference and reducing generality (but see wheat study)
Meta-analyses have been the standard solution to this problem, but precision yield data holds enormous promise.
However, its use is limited for several reasons: 1) Lack of standardized formats between equipment types, 2) Sensor calibration required for field-level accuracy, and 3) unfamiliarity with spatial statistics.
Ecosystem services can influence both the mean and variability of yield in agroecosystems (@redhead2020).
Typically only averages (means) are considered, but higher stability (lower variance) in yield can also be valuable.
There are few studies of yield variability (only those with large datasets), but precision yield data opens up the possibility of modeling within-field variability, as well as average yield.

In this paper we ask:
  1. How does crop yield change with distance from the edge of field?
  2. Does this depend on type of field edge? 
  3. Is there an intermediate distance where yield is maximized or variance is minimized?

@@

Precision yield data were collected directly from farmers across Alberta.
Farmers were solicited for yield data through local agronomists, and we received `r nrow(datSource)` field-years of data from `r length(unique(datSource$grower))` growers across a total of `r length(unique(datSource$year))` years (`r min(unique(datSource$year))`-`r max(unique(datSource$year))`).
We converted raw data to a standard csv format using Ag Leader SMS.
`r round(100*sum(grepl('(Wheat|Canola)',datSource$crop))/nrow(datSource))`% of the crop types where either wheat (_Triticum aestivum_) or canola (_Brassica napus_), two of the most common crops in rotation in Alberta.
The remaining crop types had low replication, so we constrained our analysis to only field-years containing wheat (`r sum(datSourceFinal$crop=='Wheat')`) or canola (`r sum(datSourceFinal$crop=='Canola')`).
Individual fields contained between `r min(yearsPerField)` to `r max(yearsPerField)` years of data (mean: `r round(mean(yearsPerField),1)`), containing a total of `r round(sum(datSource$npoints)/1e6,1)` million data points.

Yield data is collected in rectangles of the same length as the data interval (distance = combine ground speed $\times$ interval, typically 1 second) and the same width as the combine header (5-7 m, adjusted to account for overlap with adjacent swaths).
We extracted the size of each polygon (m$^2$), dry yield (tonnes), and the spatial location, and the sequence of collection (1 - end of harvest)
Because of the large number of yield rectangles per field (30-800 thousand), we used the centroid of each polygon as its location, treating areal data as point data
Seeding and application rates were constant across fields, so we did not consider inputs in our analysis
We used dry yield (tonnes of seed/hectare minus recorded crop moisture) as our measure of crop yield
Precision yield data can be highly variable and is prone to extreme outliers (both positive and negative), especially when combine ground speeds are low and at the beginning of rows
Therefore, we filtered data that lay outside the 95th percentiles of yield within each field, values less than or equal to zero, and values outside of the 95th percentiles of ground speed.
Due to the large number of data at each field, we sub-sampled to 50,000 data points per field to reduce computation time.

@@

Across all fields, Standard (grassy border) field boundaries were the most common boundary type, followed by Shelterbelts and Other Crops.
<!-- Not sure if we should talk about this. Not very many fields from southern regions, so it's hard to compare. Perhaps mention in Discussion -->

```{r boundaryTypeTable, echo=FALSE, fig.cap='Percentage of field boundaries across sampled fields.', out.width='100%'}
bLengthSummary %>% #Table version
  kable(.,format='latex',caption='Mean and SD of proportion of boundary types across all fields, along with the number of fields bordered by at least some of the boundary type.',col.names = c('Type','Mean','SD','Number of Fields'),digits = c(1,4,4,1),escape = FALSE)

@@

Average canola yield tended to increase with distance from Standard and Other Crop boundaries, leveling off at about 25 m (Figure \@ref(fig:canolaPlot) top row, Other crop and Standard panels); yield was approximately 0.1 T/ha (1.5 bu/ac) lower at the edge for both of these boundary types.
Interestingly, Shelterbelts were well-represented in the data (Table \@ref(tab:boundaryTypeTable)) but the negative edge effect was not strong (Figure \@ref(fig:canolaPlot) top row, Shelterbelt panel), indicating that they may provide better micro-climates for crops located near them.
Other boundary types were less common in the data and had correspondingly weaker effects on mean yield, but all showed a small negative edge effect.
Yield variability tended to decrease with distance from the edge of the field, indicating that edge effects contribute to yield "patchiness" as well as averages.
At Standard, Bare, and Wetland field boundaries, variability decreased until about 100 m, after which variability was constant, while the pattern was less consistent in Grassland, Other Crop, and Shelterbelt.

@@

@@

- Summary:
  - At the level of individual fields, there was substantial variation in yield with distance from the field boundary, as well as with harvest sequence and spatially within each field
  - Overall, however, the effect of field boundary on yield was inconsistent and the change in mean yield tended to be small ($\sim$0.5 T/ha for canola)

>>>>

useThese <- with(datSource,use&grepl('Wheat|Canola|Peas',crop)) #Rows to use

datSourceFinal <- datSource %>% filter(useThese)
yearsPerField <- datSourceFinal %>% count(field) %>% pull(n) #How many years of data at each unique location?

#Boundary lengths
bLengths <- read.csv('../Data/boundaryLengths.csv') %>% filter(useThese)
bNames <- c('Bare','Grassland','Other Crop','Shelterbelt','Standard','Wetland')

@@

use <- with(datSource,use&modelComplete2&crop=='Wheat')
wheatMods <- lapply(paths[use],function(x) getModelInfo(x)$smooths) %>% 
  set_names(nm=datSource$filename[use]) %>% 
  bind_rows(.id='field') %>% 
  mutate(pval=as.numeric(pval)) %>% 
  group_by(Smoother) %>% 
  summarize(meanP=mean(pval),maxP=max(pval),propSig=sum(pval<0.05)/n(),meanEDF=mean(edf),meanRef=mean(Ref.df))

use <- with(datSource,use&modelComplete2&crop=='Peas')
peaMods <- lapply(paths[use],function(x) getModelInfo(x)$smooths) %>% 
  set_names(nm=datSource$filename[use]) %>% 
  bind_rows(.id='field') %>% 
  mutate(pval=as.numeric(pval)) %>% 
  group_by(Smoother) %>% 
  summarize(meanP=mean(pval),maxP=max(pval),propSig=sum(pval<0.05)/n(),meanEDF=mean(edf),meanRef=mean(Ref.df))

rm(use,paths)

@@

Intensive agricultural production has increased over the last 100 years, and agricultural land now makes up over a third of ice-free land on Earth.
This has allowed increases in human population and increased global stability in production.
However, this is not without cost, as higher-diversity non-crops are converted to lower-diversity crops, resulting in loss of habitat and overall biodiversity of non-target organisms.
Maintaining both biodiversity and production in agroecosystems represents a seldom-considered goal of conservationists and agronomists, and hold the potential for win-win scenarios.
Key to this is the preservation of semi-natural land (SNL), which represents the interface between crops and non-crops within agroecosystems.

SNL in and around crops is important for both agricultural production and conservation, as it has the potential to create better biotic and abiotic conditions [@case2019; @weninger2021].
They are habitat for mobile organisms, and can act as sources of ecosystem services such as pollination or pest control [@albrecht2020; @gardner2021], but this can vary depending on the system [@gray2014; @quinn2017]
They also can create microclimate effects that reduce extreme temperature, trap moisture, and reduce wind speed [@kort1988; @brandle2004]
Unfortunately, most of the research is concentrated in Europe, and they tend to be less-studied on other continents
In particular, North American agroecosystems have larger fields, different crop varieties, and agronomic practices, all of which could negate or confound potential effects of SNL

<!-- - SNL has conflicting effects on crop yield because of ecosystem service provision as well as negative edge effects -->
SNL may affect yields at intermediate distances, depending on the spatial scale at which ecosystem services operate.
Edge effects cause low yields at the edge of crops because of sparse or late seedling emergence, poor microclimate, and competition with weeds.
At the same time, the centre of large fields will not receive ecosystem services if they decay with distance from edges (@kowalchuk1995).
For example, pollination services from central place foragers that nest in SNL but forage in crops drops rapidly with distance.
Therefore, yield may be maximized at intermediate distances, where the ecosystem services cancel out negative edge effects (e.g. @vanVooren2017).
This suggests a "goldilocks" field size, where negative edge effects are canceled out by ecosystem services.
<!-- - Less-considered are ecosystem ``disservices", where weeds and pests  -->
<!-- - In North America, there are financial incentives to make fields large, homogeneous, and with simple geometry -->

```{r hypotheses, echo=FALSE, fig.cap='Potential yield patterns, depending on ecosystem service effects with distance. In both figures, negative environmental effects (edge effects) are shown in red, and decrease with distance from the edge of the crop, creating an increase in yield with distance. The left figure shows a decline in ecosystem services with distance, leading to an intermediate peak in yield, while the right figure shows an increase in services with distance, leading to a peak at the centre of the field (note: Scenario 2 also occurs if ecosystem services do not change with distance).', out.width='100%'}  
  include_graphics('../Figures/ExamplePlots/hypotheses.png',dpi=NA)
```

Studies of SNL effects on crop yield also suffer from limited scope (e.g. few crop types) and small sample sizes, limiting inference and reducing generality (@baker2018; but see also @redhead2020)
Meta-analyses have been the standard solution to this problem (e.g. @albrecht2020), but precision yield data holds enormous promise.
However, its use is limited for several reasons, namely: 1) Lack of standardized formats between equipment types, 2) Sensor calibration required for field-level accuracy, and 3) low familiarity with spatial statistics.
Ecosystem services can influence both the mean and variability of yield in agroecosystems (@redhead2020); typically only averages (means) are considered, but higher stability (lower variance) in yield can also be valuable from the perspective of a grower.
There are few studies of yield variability (only those with large datasets), but precision yield data opens up the possibility of modeling within-field variability, as well as average yield.

In this paper we ask:
  1. How does crop yield change with distance from the edge of field?
  2. Does this depend on type of field edge? 
  3. Is there an intermediate distance where yield is maximized or variance is minimized?
To our knowledge, no other study has addressed the effect of field boundaries on productivity using precision yield, making this study highly novel.

@@

Precision yield data were collected directly from farmers across Alberta.
Farmers were solicited for yield data through local agronomists, and we received `r nrow(datSource)` field-years of data from `r length(unique(datSource$grower))` growers across a total of `r length(unique(datSource$year))` years (`r min(unique(datSource$year))`-`r max(unique(datSource$year))`).
We converted raw data to a standard csv format using Ag Leader SMS.
`r round(100*sum(grepl('(Wheat|Canola)',datSource$crop))/nrow(datSource))`% of the crop types where either wheat (_Triticum aestivum_) or canola (_Brassica napus_), two of the most common crops in rotation in Alberta.
The remaining crop types had low replication, so we constrained our analysis to only field-years containing wheat (`r sum(datSourceFinal$crop=='Wheat')`), canola (`r sum(datSourceFinal$crop=='Canola')`), or peas (`r sum(datSourceFinal$crop=='Peas')`).
Individual fields contained between `r min(yearsPerField)` and `r max(yearsPerField)` years of data (mean: `r round(mean(yearsPerField),1)`), containing a total of `r round(sum(datSourceFinal$npoints)/1e6,1)` million data points.

Yield data was collected in rectangles of the same length as the data interval (distance = combine ground speed $\times$ interval, typically 1 second) and the same width as the combine header (5-7 m, adjusted to account for overlap with adjacent swaths).
We extracted the size of each polygon (m$^2$), dry yield (tonnes), and the spatial location, and the sequence of collection (1 - end of harvest)
Because of the large number of yield rectangles per field (30-800 thousand), we used the centroid of each polygon as its location, treating areal data as point data
Seeding and application rates were constant across fields, so we did not consider inputs in our analysis
We used dry yield (tonnes of seed/hectare minus recorded crop moisture) as our measure of crop yield
Precision yield data can be highly variable and is prone to extreme outliers (both positive and negative), especially when combine ground speeds are low and at the beginning of rows
Therefore, we filtered data that lay outside the 95th percentiles of yield within each field, values less than or equal to zero, and values outside of the 95th percentiles of ground speed.
Due to the large number of data at each field, we sub-sampled to 50,000 data points per field to reduce computation time.

@@

Across all fields, Standard (grassy border) field boundaries were the most common boundary type, followed by Shelterbelts and Other Crops (Figure \@ref(fig:boundaryTypes)).
<!-- Not sure if we should talk about this. Not very many fields from southern regions, so it's hard to compare. Perhaps mention in Discussion -->

```{r boundaryTypes, echo=FALSE, fig.cap='Percentage of field boundaries across sampled fields.', out.width='100%'}
bLengths %>% 
  mutate(croptype=datSourceFinal$crop) %>% 
  mutate(location=ifelse(grepl('(Alvin|Dean)',filename),'Southern Alberta','Central Alberta')) %>%
  group_by(location) %>% mutate(nFields=length(unique(filename))) %>% ungroup() %>% 
  mutate(location=paste0(location,' (N field-years = ',nFields,')')) %>%
  select(filename,croptype,location,contains('prop_')) %>% 
  pivot_longer(contains('prop_')) %>% 
  mutate(name=gsub('prop_','',name),value=value*100) %>% na.omit() %>%
  mutate(name=factor(name,labels=c('Bare','Grassland','Other\nCrop','Shelterbelt','Standard','Wetland'))) %>% 
  ggplot(aes(x=name,y=value,fill=croptype))+geom_boxplot()+
  facet_wrap(~location,ncol=1) +
  labs(y='Percentage of field boundary',fill='Crop')+
  scale_fill_manual(values=c('grey50','grey80','white'))+
  theme(axis.title.x = element_blank(),legend.position = 'bottom')

# bLengthSummary %>% #Table version
#   kable(.,format='latex',caption='Mean and SD of proportion of boundary types across all fields, along with the number of fields bordered by at least some of the boundary type.',col.names = c('Type','Mean','SD','Number of Fields'),digits = c(1,4,4,1),escape = FALSE)

@@

Average canola yield tended to increase with distance from Standard and Other Crop boundaries, leveling off at about 25 m (Figure \@ref(fig:canolaPlot) top row, Other crop and Standard panels); yield was approximately 0.1 T/ha (1.5 bu/ac) lower at the edge for both of these boundary types.
Interestingly, Shelterbelts were well-represented in the data (Figure \@ref(fig:boundaryTypes)) but the negative edge effect was not strong (Figure \@ref(fig:canolaPlot) top row, Shelterbelt panel), indicating that they may provide better micro-climates for crops located near them.
Other boundary types were less common in the data and had correspondingly weaker effects on mean yield, but all showed a small negative edge effect.
Yield variability tended to decrease with distance from the edge of the field, indicating that edge effects contribute to yield "patchiness" as well as averages.
At Standard, Bare, and Wetland field boundaries, variability decreased until about 100 m, after which variability was constant, while the pattern was less consistent in Grassland, Other Crop, and Shelterbelt.

@@

### Pea fields

At the field level, both yield average and yield variability (SD of yield) of yield were spatially (average p-values: `r with(peaMods,formatC(meanP[Smoother=='s(E,N)']))`, `r with(peaMods,formatC(meanP[Smoother=='s.1(E,N)'],3))`) and temporally dependent (`r with(peaMods,formatC(meanP[Smoother=='s(r)'],3))`, `r with(peaMods,formatC(meanP[Smoother=='s.1(r)'],3))`), 
There was a strong negative effect of ground speed on yield and variability of peas (Figure \@ref(fig:peasPlotSpeed)), but harvest sequence did not distort the average or SD of wheat yield in a systematic way (Figure \@ref(fig:wheatPlotSpeed), right panel).

Average pea yields increased with distance from Shelterbelt, Standard, and Wetland field boundaries.
Interestingly, the reduction at the edge was slightly higher for Wetlands, but yield increased with distance more quickly, i.e. yields were higher at distance for Wetlands.
Variability decreased with distance from Grassland and Standard boundaries

```{r peaPlot, echo=FALSE, fig.cap='Field boundary effect on pea yield, accounting for the effect of combine speed, spatial variation, and harvest sequence. Upper panel represents mean yield, while the lower panel represents yield variation (i.e. "patchiness").', out.width='100%'}  
  include_graphics('../Figures/ModelSummary3a_peas.png',dpi=NA)
```

@@

At the level of individual fields, both average and SD of yield varied with distance from the field boundary.
Both metrics also varied with ground speed and harvest sequence, and there were strong spatial relationships within each field, likely related to underlying soil or moisture conditions.
Yield tended to increase with distance from field boundaries before plateauing, and there was little evidence of intermediate boosts in yield, meaning that ecosystem services likely increased or did not vary with distance from field boundaries (Scenario 2, Figure \@ref(fig:hypotheses)).
However, the effect of field boundary types was inconsistent, and the change tended to be small, especially for wheat.
<!-- For example, average canola yield was about 0.5 T/ha lower close to Standard field edges, and yield SD was about 0.3 T/ha SD higher. -->
So, why did we find few consistent effects of boundary type on canola and wheat yield?
We consider a few potential reasons for this below:
1. Landscape influence
2. Boundary variables
3. Ecosystem disservices
4. Yearly/location variation
5. Lack of pest pressure/pollination deficit

Our analysis only examined boundary types, without considering the landscape context that fields were embedded in.
Some studies show the influence of landscape composition at larger spatial scales (e.g. Lan's paper) which could be particularly important when beneficial arthropods are highly mobile (e.g. bumblebee foraging) or disperse very broadly [@oberg2007].
There could also be interactions between landscape composition at different scales (e.g. our paper in AGEE), but this is not well-studied.
Landscape complexity [@fahrig2011] and connectivity are also associated with higher abundance of beneficial arthropods (Paul's new paper + others).
This means that field boundaries such as wetlands or shelterbelts may only be helpful if connected to a larger network of non-crop habitat (but see @fahrig2019).

Our definitions of boundary type were necessarily loose, as we used satellite imagery to manually classify them.
However, it may be that other boundary traits, such as size or plant species composition, are more important for spillover of ecosystems services into fields.
For example, larger [@lecq2018; @stasiov2020] or more functionally diverse (@bonifacio2011; @boughey2011) hedgerows tend to support more biodiversity than smaller hedgerows.
Increased litter cover and soil compaction increased pollen beetle predators [@sutter2018b].

Finally, pest pressure or pollination deficit may be at low enough levels that ecosystem service provision does not significantly change yield.
Canola is largely self-pollinating (at least in hybrid canola varieties, @lindstrom2016) and wheat is wind-pollinated, meaning that slight increases in pollination services may not change yields significantly in either crop [@woodcock2016].
Pest pressures may also limited by growers' application of insecticides (seed-treated or on the crop directly), and insect pest outbreaks tend to be temporally and spatially spotty.

The net flow of ecosystem services may not be positive, if beneficial organisms do not disperse from the SNL, or if pests disperse along with them.
Ecosystem disservices such as pests and weeds can also disperse from field boundaries, which may negate any positive effect of ecosystem services (@zhang2007).
@gray2014 showed that remnant forest fragments do not provide significant pest control services , but also do not depress yield.
However, this is still poorly studied, and parsing out services from disservices is impossible using crop yield data _alone_.

Since our sampling represents a limited amount of spatial and temporal coverage, there could be additional confounding variables that obscure signals of yield.
Most fields were located in Central Alberta (Dry Mixedwood region, @ABRegions2006), where boundaries contained a higher proportion of Shelterbelts (Figure \@ref(fig:boundaryTypes)), but differences from Southern AB (Foothills Fescue and Mixedgrass region) may be confounding the signal.
Between-year effects may be important, especially during dry years [@kort1988], but this was not considered in our analysis.
Future work could examine whether boundaries have differing effects during hot or cool years.

We used harvest time as smoothed variable to deal with temporal autocorrelation in the data, but this is a crude way to deal with this.
This is because in-field combine driving patterns _almost always_ induce correlation between spatial and temporal measurements (e.g. northern section of the field was harvested before the southern section), meaning that temporal and spatial random effects are difficult to separate.
There are few solutions to this using harvest data as-is, other than to ignore temporal random effects entirely, thereby constraining all random variation to be spatial.
However, it is possible to remove correlation between spatial and temporal measurements by using specially designed harvest patterns within a field (harvesting alternating strips sequentially across the field, then harvesting the remaining strips on a second pass).
Future studies could use this to estimate drift in combine sensor measurements over time.
